---
title: "hidecan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hidecan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tibble)
library(dplyr)
library(purrr)
library(stringr)
```

```{r setup}
library(hidecan)
```

## Input data

The `hidecan` package takes as input tibbles (data-frames) of GWAS and DE results or candidate genes. The input data-frames should contain some mandatory columns, depending on the type of data.

A list of example input datasets can be obtained via the `get_example_data()` function:

```{r get-example-data}
x <- get_example_data()

str(x, max.level = 1)
```

### GWAS results

GWAS results should be provided as a tibble or data-frame, with one row per genetic marker. The data-frame should contain at least the following columns:

-   `chromosome`: character column, giving the ID of the chromosome on which each marker is located;

-   `position`: numeric column, the physical position along the chromosome (in base pairs - bp) of the marker;

-   either `score` or `padj`: numeric column, providing either the score (i.e. `-log10(p-value)`) or the adjusted p-value of the marker. If a `score` column is provided, the `padj` column will be ignored. If only a `padj` column is provided, a `score` column will be constructed as `-log10(padj)`.

Any other columns present in the data-frame will be ignored. An example of valid input is shown below:

```{r example-data-gwas}
head(x[["GWAS"]])
```

### Differential expression results

DE results should be provided as a tibble or data-frame, with one row per gene. The data-frame should contain at least the following columns:

-   `chromosome`: character column, giving the ID of the chromosome on which each gene is located;

-   `start` and `end`: numeric columns, giving the starting and end position of the gene along the chromosome (in bp). These two columns will be used to calculate the position of the gene (column `position`) as the half-way point between the start and end of the gene.

-   either `score` or `padj`: numeric column, providing either the score (i.e. `-log10(p-value)`) or the adjusted p-value of the gene. If a `score` column is provided, the `padj` column will be ignored. If only a `padj` column is provided, a `score` column will be constructed as `-log10(padj)`.

-   either `foldChange` or `log2FoldChange`: numeric column, giving either the fold-change or log2(fold-change) of the gene. If a `log2FoldChange` column is provided, the `foldChange` column will be ignored. If only a `foldChange` column is provided, a `log2FoldChange` will be constructed as `log2(foldChange)`.

Any other columns present in the data-frame will be ignored. An example of valid input is shown below:

```{r example-data-de}
head(x[["DE"]])
```

(Note that in the example dataset, some genes have missing values in the `padj` column; this corresponds to genes that have been filtered out via [independent filtering in the DEseq2 package](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pvaluesNA)).

### Candidate genes

A list of candidate genes (e.g. genes previously found associated with a trait of interest based on literature search) can be provided as a tibble or data-frame, with one row per gene. This data-frame can also contain variants of interest (see below). The data-frame should contain at least the following columns:

-   `chromosome`: character column, giving the ID of the chromosome on which each gene is located;

-   `start` and `end`: numeric columns, giving the starting and end position of the gene along the chromosome (in bp). These two columns will be used to calculate the position of the gene (column `position`) as the half-way point between the start and end of the gene. To specify a genomic variant, simply set the `start` and `end` columns to the physical position of the variant.

-   `name`: character column, giving the name of the candidate gene that will be displayed in the HIDECAN plot. Set to `NA` for any subset of genes to remove their label in the plot (can help if many genes are very close, to avoid cluttering the plot).

Any other columns present in the data-frame will be ignored. An example of valid input is shown below:

```{r example-data-can}
head(x[["CAN"]])
```

## Creating a HIDECAN plot

The `hidecan_plot()` function is the main function that generates the HIDECAN plots. It takes as input the data-frames presented above, as well as the score and log2(fold-change) thresholds used to select the significant markers and genes.

```{r hidecan-plot, fig.width = 10, fig.height = 10}
hidecan_plot(
  gwas_list = x[["GWAS"]],          
  de_list = x[["DE"]],              
  can_list = x[["CAN"]],            
  score_thr_gwas = -log10(0.0001),  ## sign. threshold for GWAS
  score_thr_de = -log10(0.05),      ## sign. threshold for DE
  log2fc_thr = 0                    ## log2FC threshold for DE
)
```

Note that it is possible to provide only a subset of the possible input data, e.g. only GWAS results and a list of candidate genes:

```{r hidecan-plot-gwas-can-only, fig.width = 10, fig.height = 10}
hidecan_plot(
  gwas_list = x[["GWAS"]],          
  can_list = x[["CAN"]],            
  score_thr_gwas = -log10(0.0001)
)
```

## Colour genes by log2(fold-change)

By default, in a HIDECAN plot, the points representing both significant markers and DE genes are coloured according to their GWAS/DE score. However, it is possible to colour the DE genes by their log2(fold-change) value instead, by setting the `colour_genes_by_score` argument of the `hidecan_plot()` function to `FALSE`:

```{r hidecan-genes-colour-log2fc, fig.width = 10, fig.height = 10}
hidecan_plot(
  gwas_list = x[["GWAS"]],          
  de_list = x[["DE"]],              
  can_list = x[["CAN"]],            
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.05),
  log2fc_thr = 0,
  colour_genes_by_score = FALSE
)
```

Genes with a negative log2(fold-change) will be represented with a shade of blue, and genes with a positive log2(fold-change) will be represented with a shade of red.

## More than one GWAS, DE or candidate gene list

The `hidecan_plot()` function can take as an input lists of data-frames for GWAS results, DE results or candidate genes. This way, it is possible to visualise more than one GWAS or DE analyses at once, for example if investigating several traits at once or comparing more than two treatment groups.

For this example, we'll focus on chromosomes 7 and 8.

```{r multiple-gwas-input, fig.width = 10, fig.height = 3}
library(dplyr)
library(purrr)
library(stringr)

## Retaining only markers and genes on chromosomes 7 and 8
x_small <- x |> 
  map(~ filter(.x, str_detect(chromosome, "(07|08)")))

## Creating a second GWAS result tibble by shuffling 
## the marker scores from the original data
gwas_1 <- x_small[["GWAS"]]
gwas_2 <- gwas_1 |> 
  mutate(score = sample(score))

hidecan_plot(
  gwas_list = list(gwas_1, gwas_2),
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.05),
  log2fc_thr = 0
)
```

By default, the two GWAS tracks on the y-axis will be given unique names, as can be seen above. It is possible to customise this by naming the elements in the input list:

```{r multiple-gwas-input-named, fig.width = 10, fig.height = 3}
hidecan_plot(
  gwas_list = list("Trait 1" = gwas_1, 
                   "Trait 2" = gwas_2),
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.05),
  log2fc_thr = 0
)
```

## Defining chromosomes length

By default, the `hidecan_plot()` function calculates the length of the different chromosomes based on the input data, by looking at the maximum position of genes and markers in each chromosome. However, it is also possible to pass on a tibble of chromosome length (in bp) through the `chrom_length` argument.

```{r show-chrom-length, fig.width = 10, fig.height = 10}
library(tibble)

## Tibble of chromosomes length as recorded in Ensembl Plants
potato_chrom_length <- c(
  ST4.03ch00 = 45813526,
  ST4.03ch01 = 88663952,
  ST4.03ch02 = 48614681,
  ST4.03ch03 = 62190286,
  ST4.03ch04 = 72208621,
  ST4.03ch05 = 52070158,
  ST4.03ch06 = 59532096,
  ST4.03ch07 = 56760843,
  ST4.03ch08 = 56938457,
  ST4.03ch09 = 61540751,
  ST4.03ch10 = 59756223,
  ST4.03ch11 = 45475667,
  ST4.03ch12 = 61165649
) |> 
  enframe(name = "chromosome",
          value = "length")

hidecan_plot(
  gwas_list = x[["GWAS"]],          
  de_list = x[["DE"]],              
  can_list = x[["CAN"]],
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.05),
  log2fc_thr = 0,
  chrom_length = potato_chrom_length
)
```

Note that in this case the difference is not huge.

## Controlling the plot properties

The `hidecan_plot()` function offers several arguments to control different aspects of the HIDECAN plot. For example, it is possible to specify the number of rows or columns the plot should have, through the `n_rows` and `n_cols` arguments. Note that only one of these arguments will be considered (`n_rows` takes precedence):

```{r hidecan-nrows, fig.width = 10, fig.height = 8}
## Specifying the number of rows
hidecan_plot(
  gwas_list = x[["GWAS"]],          
  de_list = x[["DE"]],              
  can_list = x[["CAN"]],            
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.05),
  log2fc_thr = 0,
  n_rows = 3
)
```

```{r hidecan-ncols, fig.width = 10, fig.height = 10}
## Specifying the number of columns
hidecan_plot(
  gwas_list = x[["GWAS"]],          
  de_list = x[["DE"]],              
  can_list = x[["CAN"]],            
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.005),
  log2fc_thr = 0,
  n_cols = 3
)
```

In addition, it is possible to:

-   add a title and subtitle to the plot (`title` and `subtitle` arguments);

-   control the position of the legend (`legend_position` argument);

-   control the size of the points (`point_size` argument);

-   control the size of the labels and padding around the text for the candidate genes (`label_size` and `label_padding` arguments).

## Viewport error

If you are working on RStudio, you may encounter the following error:

```{r show-window-size-error, echo = FALSE, error = TRUE, fig.width = 0, fig.height = 0}
hidecan_plot(gwas_list = x[["GWAS"]],
             de_list = x[["DE"]],
             can_list = x[["CAN"]],
             score_thr_gwas = -log10(0.0001),
             score_thr_de = -log10(0.005),
             log2fc_thr = 0,
             label_size = 2)
```

This is caused by the plotting window being too small. Try increasing the size of the plotting window on the RStudio console. Alternatively, you can save the plot into an object, then use `ggplot2::ggsave()` to save it into a file:

```{r use-ggsave, eval = FALSE}
p <- hidecan_plot(
  gwas_list = x[["GWAS"]],
  de_list = x[["DE"]],
  can_list = x[["CAN"]],
  score_thr_gwas = -log10(0.0001),
  score_thr_de = -log10(0.005),
  log2fc_thr = 0,
  label_size = 2
)

ggplot2::ggsave("hidecan_plot.pdf", p, width = 10, height = 10)
```
